---
title: "Species Distribution Modeling (SDM) Pipeline Guide"
subtitle: "Configuration, Execution via `runMe.R`, and Parameter Reference"
author: "T. Micheletti & L. Hildebrand"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    theme: cosmo
    highlight: tango
    number_sections: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  eval = FALSE,
  comment = "#>"
)
```

# Overview

This pipeline runs a multi-scale species distribution modeling (SDM) framework built on SpaDES and Boosted Regression Trees (`dismo::gbm.step`). It standardizes predictions across three geographic and operational scales:

1. **European Climate Scale** (~50 km, CHELSA bioclimatic predictors, EBBA2 occurrence data).
2. **German Landscape Scale** (~1000 m, Thünen land-use/cover covariates, DDA / MhB data).
3. **German Habitat Scale** (~200 m, fine-scale Thünen habitat covariates, MhB point counts).

The pipeline is driven **exclusively** by three configuration CSV files (under `data/`), eliminating hardcoded model settings in module scripts.

---

# Phase 1: The Three Config Files

All run-level options must be set in the 3 CSV files under `data/` (moved out of the project root 2026-09-28).

```text
project_root/
└── data/
    ├── speciesCanonical.csv          # Species inclusion and canonical naming
    ├── speciesConfig_general.csv     # Per-species scales, resolutions, modes, & BRT settings
    └── speciesConfig_predictors.csv  # Exact candidate predictors for "table" mode
```

## `speciesCanonical.csv`

Governs **which species run** and maps canonical scientific and vernacular names.

| Column | Description | Example / Values |
|:---|:---|:---|
| `latin_name` | Scientific name matching raw occurrence records | `"Lanius collurio"`, `"Vanellus vanellus"` |
| `german_name` | Canonical German vernacular name | `"Neuntöter"`, `"Kiebitz"` |
| `english_name` | Canonical English vernacular name | `"Red-backed Shrike"`, `"Northern Lapwing"` |
| `euring_code` | Official EURING bird identifier code | `15150` |
| `include` | **Master Run Switch:** Set `"X"` to run; leave blank `""` to omit | `"X"` or `""` |

> **Rule:** Only species with `include == "X"` are loaded into `sharedSpecies`. Do not hardcode species subsets in `runMe.R`.

---

## `speciesConfig_general.csv`

Governs **per-species and per-scale operational parameters**. Each species contains up to 3 rows corresponding to its modeling scales (`climate`, `landscape`, `habitat`).

| Column | Description | Valid Values & Options |
|:---|:---|:---|
| `species` | Scientific name matching `speciesCanonical.csv` | `"Lanius collurio"` |
| `scale` | Operational modeling tier | `"climate"`, `"landscape"`, `"habitat"` |
| `resolution_m` | Raster cell resolution in meters | `50000` (climate), `1000` (landscape), `200` (habitat) |
| `data_source` | Survey data source | `"EBBA2/CHELSA"`, `"DDA territories"`, `"MhB point counts"` |
| `thinning_dist_m` | Minimum spatial thinning distance in meters | `0` (none), `500`, `1000` |
| `brt_start_lr` | Starting learning rate for BRT step optimization | `0.01` (coarse/broad), `0.05`, `0.08` (fine-scale) |
| `brutzeitcode_filter` | Optional breeding code filter threshold | `"C"`, `"B"`, or blank |
| `predictor_mode` | **Predictor resolution mechanism:**<br>• `"table"`: Use exact list from `speciesConfig_predictors.csv`<br>• `"all"`: Use all valid environmental layers<br>• `"auto"`: Run block-CV collinearity selection (`select07Blockcv`) | `"table"`, `"all"`, `"auto"` |
| `spatial_term` | Opt-in spatial coordinates: appends projected `x` and `y` to the candidate tree splits | `"X"` (enable) or blank `""` (disable) |

---

## `speciesConfig_predictors.csv`

Consulted whenever `predictor_mode == "table"`. It lists the candidate predictors per species across scales.

```csv
species,climate,landscape,habitat
"Lanius collurio","","hedges, grassland, fallow","hedges, grassland, built_up"
"Vanellus vanellus","","grassland, summer_cereals","grassland, root_crops"
```

### The Hedges Rule
- If a species should use hedges, write `hedges` in the `landscape` or `habitat` column.
- The pipeline automatically loads the backfilled Thünen hedge layer.
- If `hedges` is omitted from the cell, the species will not use it.
- **Typo Protection:** If a requested predictor name is misspelled (e.g., `"hedge"` instead of `"hedges"`), the pipeline halts immediately with an explicit error indicating the missing column name.

---

# Phase 2: Configuring `runMe.R`

`runMe.R` serves as the master driver script for the SpaDES simulation.

## 1. Dynamic Config Ingestion

Ensure the top of `runMe.R` loads parameters dynamically from the CSV files:

```{r}
source("sharedConfig.R")
source("sharedSpeciesCanonical.R")
source("sharedSpeciesConfig.R")

# 1. Load active species list from data/speciesCanonical.csv
speciesCanonical <- loadSpeciesCanonical("data/speciesCanonical.csv")
sharedSpecies    <- canonicalIncludedSpecies(speciesCanonical)

# 2. Ingest operational and predictor tables
perSpeciesGeneralConfig <- loadSpeciesGeneralConfig("data/speciesConfig_general.csv")
speciesPredictorTable   <- loadSpeciesPredictorConfig("data/speciesConfig_predictors.csv")

# 3. Extract per-species flags
predictorsToUse   <- extractPredictorMode(perSpeciesGeneralConfig)
spatialTermConfig <- extractSpatialTermSpecies(perSpeciesGeneralConfig)
```

## 2. Setting Module Parameters (`params`)

```{r}
params <- list(
  # -------------------------------------------------------------
  # Module 1: Data Preparation
  # -------------------------------------------------------------
  dataPrep_Monitor = list(
    species    = sharedSpecies,                      # Dynamic species list
    years      = 2017:2023,                          # Survey monitoring years
    dataDir    = file.path(sharedPath, "data"),      # Path to input rasters
    scratchDir = file.path(sharedPath, "scratch")    # Fast SSD scratch directory
  ),

  # -------------------------------------------------------------
  # Module 2: Collinearity & Model-Ready Inputs
  # -------------------------------------------------------------
  inputs_Monitor = list(
    species               = sharedSpecies,
    predictorsToUse       = predictorsToUse,         # "table", "all", or "auto"
    speciesPredictorTable = speciesPredictorTable,   # Explicit lists per species
    spatialTermConfig     = spatialTermConfig,       # TRUE/FALSE spatial term flags
    
    # Scale resolutions (in meters)
    climateResolutionM    = 50000,
    landscapeResolutionM  = 1000,
    habitatResolutionM    = 200,
    
    # Collinearity parameters (used when predictor_mode == "auto")
    threshold             = 0.7,                     # Spearman |r| pruning threshold
    univar                = "gam",                   # Univariate model in select07Blockcv
    
    # Spatial cross-validation blocking
    runSpatialBlocking    = FALSE,                   # Set TRUE to calculate spatial folds
    kFolds                = 5                        # Number of spatial CV folds
  ),

  # -------------------------------------------------------------
  # Module 3: Model Fitting (BRT)
  # -------------------------------------------------------------
  models_Monitor = list(
    species                 = sharedSpecies,
    perSpeciesGeneralConfig = perSpeciesGeneralConfig, # Slices brt_start_lr per species
    treeComplexity          = 3,                       # Interaction depth
    bagFraction             = 0.75,                    # Subsampling fraction
    maxTrees                = 10000,                   # Maximum tree iterations
    tolerance               = "fixed"                  # Convergence criterion
  ),

  # -------------------------------------------------------------
  # Module 4: Evaluation & Prediction
  # -------------------------------------------------------------
  evaluation_Monitor = list(
    species         = sharedSpecies,
    metrics         = c("AUC", "TSS", "Deviance"),
    savePredictions = TRUE,
    outputDir       = file.path(sharedPath, "outputs")
  )
)
```

---

# Phase 3: Launching the Workflow

## Interactive Execution (RStudio / R GUI)

```{r}
# 1. Verify working directory is project root
getwd()

# 2. Source the driver script
source("runMe.R")
```

## Batch / High-Performance Computing (CLI)

For headless nodes or cluster environments, run in background with logging:

```bash
cd /path/to/SDM_Monitoring
nohup Rscript runMe.R > runMe_$(date +%Y%m%d_%H%M%S).log 2>&1 &
```

Monitor run output:
```bash
tail -f runMe_*.log
```

---

# Phase 4: Workflow Execution Architecture

```text
[speciesCanonical.csv]          ──> sharedSpecies
[speciesConfig_general.csv]     ──> predictor_mode, spatial_term, brt_start_lr
[speciesConfig_predictors.csv]  ──> candidate predictor table
         │
         ▼
 1. dataPrep_Monitor
    ├── Crop / cache CHELSA climate rasters (50 km)
    ├── Load / backfill Thünen land-use & hedges layers (1000 m & 200 m)
    └── Pool presence/absence occurrences per species
         │
         ▼
 2. inputs_Monitor
    ├── Write diagnostic Spearman correlation plots (.png)
    ├── Resolve predictors:
    │     • "table" ──> Validate & select from speciesConfig_predictors.csv
    │     • "all"   ──> Select all environmental candidate columns
    │     • "auto"  ──> Prune via select07Blockcv()
    ├── Append "x" and "y" if spatial_term == "X"
    └── Emit model-ready data frames:
          ID + occurrence + foldID + predictors
         │
         ▼
 3. models_Monitor
    ├── Retrieve brt_start_lr from perSpeciesGeneralConfig
    ├── Fit Boosted Regression Trees (dismo::gbm.step)
    └── Optimize tree depth & learning rate
         │
         ▼
 4. evaluation_Monitor
    ├── Calculate CV metrics (AUC, TSS, deviance explained)
    └── Export spatial prediction rasters (.tif)
```

---

# Phase 5: Outputs and Diagnostic Validation

## 1. Correlation Plots
Generated in the directory defined by `corrplotDir` (e.g. `outputs/corrplots/`):
- `[Species]_EU.png`: European climate covariate collinearity matrix.
- `[Species]_landscape_pooled.png`: German landscape scale correlation matrix.
- `[Species]_habitat_pooled.png`: German habitat scale correlation matrix.

## 2. Terminal Log Verification
During execution, verify module output in the console:

```text
  -- Lanius_collurio --------------------------
Pooled records: 1842 (312 pres / 1530 abs)
Predictors used (5): grassland hedges built_up x y
```

- If `x` and `y` are listed, `spatial_term` was successfully applied.
- If `hedges` is listed, it was successfully validated and included.

## 3. Typo Interception (No Silent Drops)
If a predictor is misspelled in `speciesConfig_predictors.csv` (e.g. `"hedge"` instead of `"hedges"`), execution halts immediately:

```text
[PREDICTOR ERROR] Species 'Lanius collurio' [HABITAT scale]:
Requested predictor(s) "hedge" from speciesConfig_predictors.csv were NOT found in occurrence dataset!
Available data columns are: grassland, winter_cereals, maize, hedges, trees, water, elevation, slope
```

**Resolution:** Correct the name in `speciesConfig_predictors.csv` and re-run.