---
title: "birdMonitor: Species Distribution Modeling Pipeline"
subtitle: "Configuration, Execution via `runMe.R`, and Reference"
author: "T. Micheletti & L. Hildebrand"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    theme: cosmo
    highlight: tango
    number_sections: true
  github_document:
    toc: true
    toc_depth: 3
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  eval = FALSE,
  comment = "#>"
)
```

# Overview

Code to run the models for the UFZ bird monitor. This document covers how
to actually run the pipeline — locally and on the EVE cluster — which
parameters/tables to check before a run, and what each output is.

This pipeline runs a multi-scale species distribution modeling (SDM)
framework built on SpaDES and Boosted Regression Trees (`dismo::gbm.step`),
combined via a ridge-regression meta-model, for the UFZ bird monitor. It
standardizes predictions across three geographic/operational scales, then
combines them and reports on the result:

1. **European climate scale** (~50 km) — CHELSA bioclimatic predictors,
   EBBA2 occurrence data.
2. **German landscape scale** (~1 km) — Thünen land-use/cover covariates,
   DDA territories (or MhB point counts, for specific species).
3. **German habitat scale** (~200 m) — fine-scale Thünen habitat
   covariates, MhB point counts.
4. **Meta-model** — combines the three scales' suitability predictions via
   ridge regression, trained on real habitat occurrence data.
5. **Multi-species index/report** — annual report (per-species and
   combined indices, change maps) and regional gridded index maps, built
   from the meta-model's output.

Code is organized as 4 SpaDES modules (`dataPrep_Monitor` ->
`inputs_Monitor` -> `models_Monitor` -> `runIndex_Monitor`, one GitHub
repo/submodule each), driven by `runMe.R`. Model settings are driven by 3
configuration CSV files (under `data/`) rather than hardcoded in module
scripts.

For **why** things are built the way they are, see:

- **`DECISIONS.md`** — every methodology/model decision (data sources,
  thresholds, predictor selection), dated, with rationale.
- **`improvements.md`** — methodological extensions still under discussion,
  not yet decided.
- **`TODO.md`** — engineering/completeness gaps.

# 0. Before you run anything

**Branch.** As of 2026-09-25, everything described here (the config CSVs,
the per-species wiring, the Buteo/Star landscape routing) lives on
`feature/reconcile-with-v2-flexible-config` in the `dataPrep_Monitor`/
`inputs_Monitor`/`models_Monitor` submodules — not yet merged to `main`.
Root `birdMonitor` itself is on `feature/config-data-folder`, and
`runIndex_Monitor` (added later, its own repo) is still on `main` — the
four repos are deliberately NOT all on the same branch name right now.
`runMe.R` runs whatever is checked out on disk in each `modules/<name>/`
folder (`useGit = FALSE`), so before running, confirm each repo is
actually on the branch you expect:

```bash
cd birdMonitor && git branch --show-current
cd modules/dataPrep_Monitor && git branch --show-current
cd ../inputs_Monitor && git branch --show-current
cd ../models_Monitor && git branch --show-current
cd ../runIndex_Monitor && git branch --show-current
```

Should print, in order: `feature/config-data-folder`,
`feature/reconcile-with-v2-flexible-config` (x3), `main`.

**Packages.** `runMe.R` installs/manages its own R package library via
`Require`/`pak` (see its own `paths`/`packages` -- no separate manual
install step needed under normal use). If you ever see a `could not find
function` error partway through a run, that's a missing package Require
didn't catch — install it and re-run.

**Raw data.** Must already exist under `inputs/` before `dataPrep_Monitor`
runs (it never downloads survey data itself — only predictor rasters). The
default paths (relative to `inputs/`), all `defineParameter(..."Subpath")`
in `dataPrep_Monitor.R`:

- `response/raw/ornitho/ebba2_data_occurrence_50km.csv` + `ebba2_grid50x50_v1.shp`
- `response/raw/MhB/dbird_observations_CBBM.csv` (the raw MhB point-count
  CSV — used at habitat scale for everyone, AND at landscape scale for
  whichever species use `"MhB point counts"` in `speciesConfig_general.csv`)
- `response/raw/MhB/MhB_Probeflaechen_DE_S2637_epsg25832.shp` (routes —
  shared by habitat and landscape scale)
- `response/raw/territories/BirdStats_Daten2005-2024D_alle.xlsx` (DDA
  territories) + `BirdStats_Visits2005-2024D.xlsx` (DDA visited routes) —
  only actually read if at least one species still uses the default
  `"DDA territories"` data source at landscape scale.

# 1. Running locally (`runMe.R`)

```{r}
source("runMe.R")
```

This runs all 4 modules end to end (`dataPrep_Monitor` -> `inputs_Monitor`
-> `models_Monitor` -> `runIndex_Monitor`) via
`SpaDES.core::simInitAndSpades()`. Outputs land under `outputs/<runName>/`.

## Batch / non-interactive execution

For a long run you don't want tied to an open terminal (headless nodes,
overnight runs), launch in the background with logging:

```bash
cd /path/to/birdMonitor
nohup Rscript runMe.R > runMe_$(date +%Y%m%d_%H%M%S).log 2>&1 &
```

Monitor progress:

```bash
tail -f runMe_*.log
```

## Parameters to check/adjust before a run

- **`runNameBase`** (top of `runMe.R`) — bump this by hand for a run whose
  cached outputs shouldn't be reused (a real config change). A timestamp is
  always appended automatically (`test1_20260927_091500`), so every run
  gets its own `outputs/` folder regardless — see `DECISIONS.md` if you
  want the history of why it's built this way.
- **Species subset, for a test run** — do NOT hand-type a species vector
  anywhere in `runMe.R`; `sharedSpecies` (from `sharedConfig.R`) is the
  single source of truth every module and the cluster path both read, and
  a hand-typed copy is easy to forget to revert. Instead, temporarily mark
  just the species you want to test `"X"` in `data/speciesCanonical.csv`'s
  `include` column (and the rest not-included) — `sharedSpecies` is built
  from that column, so every module and the cluster path see the same
  restricted roster automatically, with nothing else to remember to revert.
- **`predictionYears`** (`sharedConfig.R`) — which years to generate
  PREDICTION rasters for, across every scale and the meta-model. Restrict
  this alone (e.g. to `c(2005, 2015, 2025)`) to test a small set of output
  years without shrinking any scale's training data or breaking the
  hedges-backfill mechanism — both are handled automatically (see
  `DECISIONS.md`'s 2026-09-28 "Decouple fitting years from prediction
  years" entry). `sharedHabitatYears`/`sharedLandscapeYears` are each
  scale's separate FITTING-year constraint (real survey data availability,
  not a free choice) — a `warning()` fires if either is pushed outside its
  known real-data range.

## Config tables to check before a run

Both live under `data/` (not the repo root — moved there 2026-09-28), loaded
once by `runMe.R` via `sharedSpeciesConfig.R`. Missing either file is fine —
every module falls back to its own shared defaults exactly as if these
didn't exist.

**`speciesConfig_general.csv`** — exactly 3 rows per species (climate/
landscape/habitat):

| column | wired to real effect? | what it does |
|---|---|---|
| `resolution_m` | **Yes** | per-species resolution override (e.g. Milvus milvus's coarser landscape window) — `dataPrep_Monitor` generates each scale's distinct resolutions and re-keys occurrence extraction by it; `inputs_Monitor` groups spatial blocking by it; `models_Monitor` resolves each species' own model/prediction directory from it |
| `data_source` | **Yes** (landscape rows only) | `"DDA territories"` (default) or `"MhB point counts"` — routes that species' landscape-scale occurrence construction through a different raw source entirely, see `DECISIONS.md` |
| `thinning_dist_m` | **Yes** | per-species spatial thinning distance override |
| `brutzeitcode_filter` | **Yes** (habitat rows only) | ATLAS_CODE prefix filter (e.g. `"C"` = confirmed-breeding only), on top of the existing global filter |
| `spatial_term` | **Yes** | `"X"` opts a species into the `x`/`y` spatial-trend-surface predictor (deliberately opt-in, not blanket -- see `DECISIONS.md`'s 2026-09-26 entries) |

**`speciesConfig_predictors.csv`** — many rows per species, one per
predictor, columns `species`/`climate`/`landscape`/`habitat`. The ONLY
source of a species' candidate predictors — every species must be listed
here (a species missing here is a hard error, not a silent fallback).
Toggle a predictor off just by blanking its cell for that species+scale.
Independently, `inputs_Monitor`'s `dropCollinearPredictors` parameter
(module-level, default `FALSE`) controls whether that per-species list is
further pruned for collinearity via `select07Blockcv()` — see
`DECISIONS.md`'s 2026-09-28 entry.

## Output layout

```text
outputs/<runName>/
  <climate label>/       # e.g. scale_50 -- europe/climate-scale outputs
  <habitat label>/       # e.g. scale_02 -- habitat-scale outputs
  <landscape label>/     # e.g. scale_1  -- landscape-scale outputs
  metamodel_<labels>/    # meta-model outputs (per-species/year prediction rasters) --
                         # this is what runIndex_Monitor reads as `metaDir`
  annual_report/         # runIndex_Monitor's annual report (see below)
  regional_index/        # runIndex_Monitor's regional gridded index maps (see below)
```

# 2. The multi-species index/report (`runIndex_Monitor`)

Part of the SAME `runMe.R` run (its own SpaDES module, chained last in
`loadOrder` -- see `DECISIONS.md`'s 2026-09-28 "runIndex_Monitor" entry).
No separate script to run; one `runMe.R` invocation produces data prep ->
inputs -> models -> index/report. Configure it via `runMe.R`'s
`params$runIndex_Monitor` block: `species` (the full roster its
`checkAllInputs` event waits on before computing anything -- matters mainly
for a cluster run, where different species' `metaModel()` tasks finish at
unpredictable times), `indexSpecies` (optional subset the report/index
actually covers -- leave unset to use all of `species`; an index/change-map
for a species with no `metaModel()` output is meaningless, not just empty),
`baselineYear`/`allYears`/`currentYear`/`restrictedYears`.

Before computing anything, it warns (never stops) about any specific
year/comparison it can't actually satisfy given which years'
`metaModel()` output exists on disk -- e.g. `"species X: vsLastYear needs
year 2024's metaModel() output, which is missing"`.

Produces, under `outputs/<runName>/annual_report/`:

- `species_index.csv` — per-species baseline-100 index series
- `combined_index_{sbi,analytical,msi,chain}.csv` — four combined-index
  methods (report all four together, not just one — see `computeAnnualReport()`'s
  docstring for why they can legitimately disagree at the margins)
- `combined_index_chain_restricted.csv` — Chain index restricted to
  non-hindcast years only (a robustness check)
- `SR_map_<year>.tif` — species richness map
- `change_{vsBaseline,vs5YearsAgo,vsLastYear}_{community,<species>}.tif` —
  prevalence-change/gain-loss maps

And under `outputs/<runName>/regional_index/`:
`regional_index_<N>km_{raw,smoothed}.tif` for each of
`sharedRegionalCellSizesM`'s grid sizes (10/20/50km by default).

# 3. Running on the EVE cluster

**Status as of 2026-10-02: the whole workflow is wired for EVE, but has never
been submitted to the real cluster -- expect to debug the first run.** See the
`.claude/skills/eve-cluster/` skill for onboarding constraints (VPN-only
access, module system, Transfer Queue for downloads, memory/node sizing,
fair-share priority).

The whole workflow runs on EVE as one dependency chain, submitted by a single
command from an EVE login node:

```text
prep   (dataPrep_Monitor + inputs_Monitor)      cluster/eve_prep.sbatch
  -> model arrays: europe, habitat, landscape    modules/models_Monitor/cluster/eve_array_*.sbatch
     (one task per included species; each species' meta-model
      runs inside whichever of its scale tasks finishes last)
    -> index (runIndex_Monitor)                  cluster/eve_index.sbatch
```

```bash
# on an EVE login node, from the repo root, after VPN + clone + rsync of
# inputs/ and cache/ + R packages installed (compute nodes have throttled internet):
EVE_R_MODULE=<name from `module spider R`> \
EVE_PARTITION=<partition from `sinfo -s`> \
BIRDMONITOR_RUNNAME=test4 \
  bash cluster/submit_eve_pipeline.sh
```

- The array size is `--array=1-N` where N is the number of `include = "X"` rows
  in `data/speciesCanonical.csv` (currently 11, Milvus milvus excluded); the
  task index maps to that species' position among the included rows. Keep the
  four `.sbatch` files in sync when the roster changes.
- `runMe.R` runs a single stage when `BIRDMONITOR_STAGE` is set (`prep` or
  `index`; default `all` = everything in one session, the local behaviour --
  see `tools/sharedStageConfig.R`). `BIRDMONITOR_RUNNAME` pins the run folder
  so every stage and the model arrays write to the same `outputs/<runName>/`.
  `BIRDMONITOR_SKIP_INSTALL=1` skips the package-install block at the top of
  `runMe.R` (set automatically in the cluster jobs).
- `tools/runClusterTask.R` is the standalone entry point for one species x
  scale/meta task. It reads the `_inputs.rds`/`_predictors.rds` files that
  `inputs_Monitor` writes, which is why the prep stage must finish first.
  It builds its own `resolutionConfig`/per-species fitting years from
  `sharedConfig.R`/`speciesConfig_general.csv` (same as `runMe.R`), so a
  cluster task can never silently disagree with a full run.
- Resource requests in `eve_prep.sbatch`/`eve_index.sbatch` are unmeasured
  starting guesses; tighten them from `sacct` after the first run.
- Nothing in these jobs downloads data: put `inputs/` and `cache/` on EVE first.

```bash
# one model task, locally, for testing:
Rscript tools/runClusterTask.R --scale habitat --index 3 --run-name test4
```

# 4. Workflow architecture

```text
data/speciesCanonical.csv          --> sharedSpecies (species with include == "X")
data/speciesConfig_general.csv     --> resolution_m, thinning_dist_m, data_source,
                                        brutzeitcode_filter, spatial_term
data/speciesConfig_predictors.csv  --> the ONLY source of a species' candidate predictors
         |
         v
 1. dataPrep_Monitor
    +-- Crop/cache CHELSA climate rasters; DEM derivatives; land use/cover
    |   (each generated once per distinct resolution actually needed, not
    |   once per species -- see DECISIONS.md's 2026-09-28 entry)
    +-- Pool presence/absence occurrence data per species per scale
    |   (habitat: MhB point counts; landscape: DDA territories, or MhB-
    |   routed for specific species; climate: EBBA2)
    +-- Every covariate raster filename also encodes its own resolution
        (a second safety layer beyond the containing scale_X folder)
         |
         v
 2. inputs_Monitor
    +-- Pool per-species-year occurrence + covariates into one table per scale
    +-- Spatial blocking (real block-CV or mimicked), grouped by each
    |   species' own resolved resolution
    +-- collinearityCheck: speciesConfig_predictors.csv's table is the
    |   ONLY predictor source (hard error if a species is missing);
    |   dropCollinearPredictors optionally prunes it via select07Blockcv()
    +-- Append x/y if spatial_term == "X" for that species
    +-- Emit model-ready tables: occurrence + foldID + resolved predictors
         |
         v
 3. models_Monitor
    +-- Fit BRTs (dismo::gbm.step) per scale, per species, onto predictionYears
    |   (auto-covering habitatYears too, so the meta-model always has what
    |   it needs to train on -- see resolveScalePredictionYears())
    +-- Combine via ridge-regression meta-model (metaModel()), trained on
        real habitat occurrence (habitatYears)
         |
         v
 4. runIndex_Monitor
    +-- Waits (real-time polling in cluster mode) for every species'
    |   metaModel() output to be ready
    +-- Multi-species annual report (per-species + 4 combined-index
    |   methods, species-richness map, change maps) and regional gridded
    |   index maps
```

# 5. Output/diagnostic verification

## Correlation plots

Generated in the directory given by `inputs_Monitor`'s `corrplotDir`
parameter:

- `[Species]_EU.png` — European climate covariate collinearity matrix
  (always written).
- `[Species]_habitat_pooled.png` — German habitat-scale correlation matrix
  (written only when there are >= 2 candidate predictors, and skipped if
  already present).
- `[Species]_landscape_pooled.png` — German landscape-scale correlation
  matrix (same guard as habitat).

## Terminal log verification

During `inputs_Monitor`'s `collinearityCheck` event, verify each species'
resolved predictor list in the console:

```text
Predictors used (4): hedges, forest_cover, x, y
```

- If `x`/`y` are listed, that species' `spatial_term` was successfully
  applied.
- Whatever's listed here is EXACTLY what came from
  `speciesConfig_predictors.csv` (optionally pruned by
  `dropCollinearPredictors`) -- there is no other source.

## Predictor typo interception (no silent drops)

If a predictor is misspelled in `speciesConfig_predictors.csv` (e.g.
`"hedge"` instead of `"hedges"`), the run stops immediately, per
species+scale, before any model fitting:

```text
[PREDICTOR ERROR] Species 'Lanius collurio' [HABITAT scale]:
Requested predictor(s) "hedge" were NOT found in occurrence data!
Available columns: grassland, winter_cereals, maize, hedges, trees, water, elevation, slope, ...
```

**Resolution:** correct the name in `speciesConfig_predictors.csv` and
re-run.
